Learning CNN filters from user-drawn image makers for coconut-tree image classification

### 2020-06-11 main.tex
Updates from Overleaf

\begin{abstract}
 Identifying species of trees in aerial images is essential for land-use classification, plantation monitoring, and impact assessment of natural disasters. The manual identification of trees in aerial images is tedious, costly, and error-prone, so automatic classification methods are necessary. Convolutional Neural Network (CNN) models have well succeeded in image classification applications from different domains. However, CNN models usually require intensive manual annotation to create large training sets. A CNN may be conceptually divided into convolutional layers for feature extraction and fully connected layers for classification. We present a method that needs a minimal set of user-selected images to train the CNN's feature extractor, considerably reducing the number of images to train its classifier. The method learns the filters of each convolutional layer from user-drawn markers in image regions that discriminate classes, allowing direct human expert interference in the training process. It does not rely on optimization based on backpropagation, which considerably reduces the number of required images. We demonstrate its advantages on binary classification of coconut-tree images against a popular CNN model.
\end{abstract}

\section{Introduction}
Deep learning has proven to be applicable to different tasks, from image classification to data synthesis \cite{goodfellow2016deep}. In remote sensing, the applications have involved segmentation of terrain images \cite{kemker2018algorithms, kampffmeyer2016semantic, hamaguchi2018effective}, building identification \cite{xu2018building, lu2018detecting, liu2018multilevel}, and deforestation monitoring \cite{bragilevsky2017deep}. In this work, we are interested in 
 identifying species of trees from aerial images. The topic is important for land-use classification, plantation monitoring, and damage assessment of natural disasters. As the plantations can span very large areas, the manual identification of each tree is costly, tedious, and error-prone, and so automatic classification methods are necessary.

Classification of tree species in aerial images has been actively investigated~\cite{fassnacht2016review}. In \cite{puttemans2018comparing} and~\cite{aparna2018cnn}, the authors present automatic solutions to detect coconut trees based on convolutional neural network (CNN) models. Despite these recent advances, CNN models usually require considerable human effort in image annotation to create large training sets. Vargas-Muñoz et al.~\cite{8899005} propose an approach to mitigate the problem based on active learning. The method explores data projection techniques to allow simultaneous annotation of multiple patches as having or not a coconut tree. It then uses a CNN to identify the candidate patches with a coconut tree. We adopt another alternative in this work -- the design of simplified CNN models from reduced annotated image sets and assistance of a human expert.

The design of CNN models without a human expert as part of the training loop leaves several questions unanswered: (1) How to choose the most reasonable model, with acceptable effectiveness, and as simple as needed for a given classification problem? (2) How to train that model from a reduced number of images?  (3) Can the user explain the decisions of that model? (4) Can that model improve from human corrections? The first question requires to explore human knowledge about the classification technique and the problem of interest. The second one requires to reduce as much as possible the human effort to train a CNN model. The third issue is related to human understanding. It may explore data visualization techniques to explain the model's decisions and guide human decisions concerning the design of the model. The fourth question is also essential during training, and it is related to human control over the process. They all lead to the importance of involving human experts (users) during the machine learning process.

In this paper, we present significant advances related to the questions (1) and (2). First, we conceptually divide a CNN into convolutional layers for feature extraction and fully connected layers for feature space reduction and classification. Each convolutional layer contains a filter bank, an activation function, and alternative operations (e.g., pooling, batch normalization).  As the number of convolutional layers increases (deeper is the model), higher is the number of annotated samples required to train the model by backpropagation.

We present a method that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, a few training images, and draws markers in regions that best discriminate the classes. We explain how the classes may appear in distinct clusters of a local feature space defined by patches as extracted from marker pixels at the input of each convolutional layer. Patch size (number of weights per filter) is also a user choice. The architecture of the network can be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}. The method uses clustering and defines the weights of the filters from the centroids of the clusters. Such filters must activate the discriminant regions in the output of the given convolutional layer, and the process repeats in a layer-by-layer fashion with usually no need for new markers.

We demonstrate the advantages of this new technique on binary image classification of coconut-tree images against a popular CNN model, VGG-16~\cite{simonyan2014very}. First, the user can directly interfere and verify the markers' effectiveness in the training process.  The strategy allows user control and a better understanding of CNN models. Second, very few images per class (e.g., less than five) are usually enough. Third, by eliminating the need for optimization of convolutional layers by backpropagation, the method can considerably reduce the number of annotated images to train the fully connected layers.  Fourth, the technique is application-independent, and so it might be useful in other image classification problems.

\section{Learning filters from markers}
\label{sec:method}

Convolutional neural networks perform well in image classification tasks due to its capability of extracting features. That is, convolutional layers produce great descriptors that can be used by the classifier. Conventionally, during the training process, an algorithm tries to minimize a loss function that measures how good the classifier is performing with the features given by the feature extractor. The algorithm of optimization tries through backpropagation to find out which features to extract. Our method is based on the idea that the network designer is an expert in the application domain or knows well enough to indicate which are the relevant characteristics of the objects of interest. Hence, it does not rely on the optimization of some loss function to learn filters for those relevant characteristics. In our approach, the network designer highlights features that she believes are the ones that characterize each class by placing markers. Those markers are used to learn the filters to identify the features they represent.

Let $\D$ be a dataset of images, and let $I \in \D$ be an image with dimensions $X \times Y$. A marker $m$ is associated with a pixel $p(m) = (x, y)$ and has a label $L(m) = c$. Let $M(I)$ be the set of markers of image $I$.  For each maker $m \in M(I)$, we create a patch $P$ with dimensions $k \times k$ from a window centered in the marker's pixel as $P(I, x ,y) = I(x+i, y+j)$ for $i, j = -\floor{k/2}, \ldots, \floor{k/2}$.We pad images with $0$ so that we can create patches from border pixels. For each class, we create a set of patches made using the markers of that class. Let \[\mathcal{P}_c = \bigcup_{ I \in \D, m \in M(I) :\, L(m) = c}{P(I, p(m))}\] be the set of all patches made from markers of class $c$.

@@

\begin{table}[!t]
  \begin{center}
  \begin{tabular}{|l|c|c|c|}
  \hline
   Method & Precision & Recall & F-score \\
  \hline\hline
    I-Ours & $0.856 \pm 0.014$ & $0.836 \pm 0.020$ & $0.842 \pm 0.018$\\
    II-Ours (B) & $0.851 \pm 0.018$ & $0.834 \pm 0.023$ & $0.839 \pm 0.021$\\
    III-Ours (SVM) & $0.836 \pm 0.015 $ & $ 0.793 \pm 0.025$ & $ 0.804 \pm 0.022$\\
    IV-VGG & $0.84 \pm 0.01$ & $0.76 \pm 0.03$ & $0.77 \pm 0.02 $ \\
    V-VGG (PT) & $0.84 \pm 0.01$ & $0.75 \pm 0.03$ & $0.77 \pm 0.02 $ \\
  \hline
  \end{tabular}
  \end{center}
  \caption{Results.}
  \label{tab:results}
\end{table}

We tested three scenarios: (I) using out method to learn the feature extractor weights and the classifier with backpropagation; (II) using our method as initialization and train the whole model with backpropagation; (III) using SVM Linear as classification. For VGG-16, we tested two secenarios: (IV) initializing with Xavier initialization and training from screath; (V) using ImageNet weights and training with the training set images. The results can be seen in Table \ref{tab:results}. As the experiments were repeated three times, we reported the mean and standard deviation of each the metrics in the test set.

>>>>

\begin{abstract}
Identifying species of trees in aerial images is essential for land-use classification, plantation monitoring, and impact assessment of natural disasters. The manual identification of trees in aerial images is tedious, costly, and error-prone, so automatic classification methods are necessary. Convolutional Neural Network (CNN) models have well succeeded in image classification applications from different domains. However, CNN models usually require intensive manual annotation to create large training sets. One may conceptually divide a CNN into convolutional layers for feature extraction and fully connected layers for feature space reduction and classification. We present a method that needs a minimal set of user-selected images to train the CNN's feature extractor, reducing the number of required images to train the fully connected layers. The method learns the filters of each convolutional layer from user-drawn markers in image regions that discriminate classes, allowing better user control and understanding of the training process. It does not rely on optimization based on backpropagation, and we demonstrate its advantages on the binary classification of coconut-tree images against one of the most popular CNN models.
\end{abstract}

\section{Introduction}
Deep learning has proven to be applicable to different tasks, from image classification to data synthesis \cite{goodfellow2016deep}. In remote sensing, the applications may involve segmentation of terrain images \cite{kemker2018algorithms, kampffmeyer2016semantic, hamaguchi2018effective}, building identification \cite{xu2018building, lu2018detecting, liu2018multilevel}, and deforestation monitoring \cite{bragilevsky2017deep}, for instance. In this work, we are interested in 
 identifying species of trees from aerial images. The topic is important for land-use classification, plantation monitoring, and damage assessment of natural disasters. As the plantations can span very large areas, the manual identification of each tree is costly, tedious, and error-prone, and so automatic classification methods are necessary.

Classification of tree species in aerial images has been actively investigated~\cite{fassnacht2016review}. In \cite{puttemans2018comparing} and~\cite{aparna2018cnn}, the authors present automatic solutions to detect coconut trees based on convolutional neural network (CNN) models. Despite these recent advances, CNN models usually require considerable human effort in image annotation to create large training sets. Vargas-Muñoz et al.~\cite{8899005} propose an active learning approach to mitigate the problem. The method explores data projection techniques to allow simultaneous annotation of multiple patches as having or not one coconut tree. It then uses a CNN to identify the candidate patches with one coconut tree. We adopt another alternative -- the design of simplified CNN models from small training sets with user interaction, being the user an expert in the design of CNN models.

The design of CNN models without a human as part of the training loop leaves several questions unanswered: (1) How to find a useful and simplified model for a given classification problem? (2) How to train that model from a minimum number of annotated images?  (3) Can the user explain the decisions of the model? (4) Can the model improve from label corrections? The first question requires human knowledge about CNN modeling and the problem of interest. The second one requires to reduce the human effort to train a CNN model. The third issue is related to human understanding. It may explore data visualization techniques to explain the model's decisions and to guide the user in the design of the model. The fourth question is also essential during training, and it is related to user control over the process. They all lead to the importance of involving an expert user in the machine learning process.
In this paper, we present significant advances related to questions (1) and (2). First, we conceptually divide a CNN into convolutional layers for feature extraction and fully connected layers for feature space reduction and classification. Each convolutional layer contains a filter bank, an activation function, and alternative operations (e.g., pooling, batch normalization).  As the number of convolutional layers increases (deeper is the model), and so higher is the number of annotated images required to train the model by backpropagation.

In order to reduce the need for large annotated training sets by exploiting the user knowledge, we present a method, called FLIM (\emph{Feature Learning from Image Markers}), that needs a minimal set of images to learn the filters of each convolutional layer. The user selects the number of convolutional layers, their filter sizes, a few training images, and draws markers in regions of the input images that best discriminate the classes. We explain how the classes may appear in distinct clusters of a local feature space defined by patches as extracted from marker pixels at the input of each convolutional layer. FLIM defines the weights of the filters as the centroids of the clusters. Such filters are expected to activate discriminant regions in the output of the given convolutional layer, and the process repeats in a layer-by-layer fashion. The architecture of the CNN could also be optimized by observing its results on a small validation set or based on data visualization techniques~\cite{rauber2016visualizing}.

We demonstrate the advantages of this interactive technique on binary image classification of coconut-tree images against  VGG-16~\cite{simonyan2014very}. First, the user can better understand and control the training process by observing the effectiveness of  marker selection.  Second, very few images per class (e.g., less than five) seem to be  enough. Third, by eliminating the optimization of the convolutional layers by backpropagation, FLIM can reduce the number of annotated images to train the fully connected layers.  Fourth, FLIM is application-independent, and so it might be useful in other image classification problems.

\section{FLIM: Feature Learning  from Image Markers}
\label{sec:method}

A well-trained CNN model should detect image regions that best discriminate classes as positive activations, improve class separation in those regions by eliminating negative activations, and aggregate the resulting activations (pooling and flattening) into a highly dimensional and sparse feature space, in which even a linear classifier should separate the classes with reasonable success. Nevertheless, it is common to assume that the classes are at least piecewise linearly separable and use fully connected layers -- a Multi-Layer Perceptron (MLP) classifier -- for feature space reduction and classification.

The training of a CNN usually rely on  weight optimization by backpropagation, which requires a higher number of annotated images as deeper is the model. In FLIM, we eliminate the need for backpropagation by finding a set ${\cal F}$ of filters that can detect discriminant regions from all classes.

The user selects a few training images $I$ to compose a very small dataset $\D$ and draws labeled markers in image regions that best discriminate the classes. We with to estimate the weights of filters $F\in {\cal F}$, such that the convolution between an image $I$ and $F$ will generate positive activations in regions whose pattern is represented by $F$.

For a problem with $c$ classes, let $\lambda(p)=i\in \{1,2,\ldots,c\}$ be the label of a marker pixel $p$ from class $i$. Let ${\cal M}_I$ be a set of marker pixels drawn in image $I$ and $P_I(p)$ be a patch of size $k\times k$ pixels and $m$ image channels, around a marker pixel $p\in {\cal M}_I$. The patch values compose a \emph{local feature vector} $\vect\left[P_I(p)\right]$ with $k\times k\times m$ values. Let ${\cal P}$ be the set of all patches around marker pixels from all images $I\in {\cal D}$, with representative examples  $\vect\left[P_I(p)\right]$ from each of the $c$ classes.  
\begin{eqnarray}
{\cal P} & = & \bigcup_{I\in {\cal D},p\in {\cal M}_I, \lambda(p)\in [1,c]} P_I(p)
\label{eq.pset}
\end{eqnarray}
The patch set ${\cal P}$ defines samples from all classes in a local feature space $\Re^{k\times k\times m}$.

The classes are usually not separated in this space but the convolution of an image $I$ with a set of filters ${\cal F}$ of sizes $k\times k\times m$ should facilitate their separation in the new space

is equivalent to computing the inner product between $\vect\left[P_I(q)\right]$ and  $\vect\left[F\right]$ for all pixels $q$ in the image $I$.

 A marker $m$ is associated with a pixel $p(m) = (x, y)$ and has a label $L(m) = c$. Let $M(I)$ be the set of markers of image $I$.  For each maker $m \in M(I)$, we create a patch $P$ with dimensions $k \times k$ from a window centered in the marker's pixel as $P(I, x ,y) = I(x+i, y+j)$ for $i, j = -\floor{k/2}, \ldots, \floor{k/2}$.We pad images with $0$ so that we can create patches from border pixels. For each class, we create a set of patches made using the markers of that class. Let \[\mathcal{P}_c = \bigcup_{ I \in \D, m \in M(I) :\, L(m) = c}{P(I, p(m))}\] be the set of all patches made from markers of class $c$.

@@

\begin{table}[!t]
  \begin{center}
  \begin{tabular}{|l|c|c|c|}
  \hline
   Method & Precision & Recall & F-score \\
  \hline\hline
      FLIM + SVM & $0.856 \pm 0.011 $ & $ 0.831 \pm 0.019$ & $ 0.838 \pm 0.017$\\ 
       VGG-FE + SVM & $0.855 \pm 0.001$ & $0.816 \pm 0.007$ & $0.826 \pm 0.006 $ \\\hline

      FLIM & $0.863 \pm 0.002$ & $0.849 \pm 0.005$ & $0.854 \pm 0.004$\\
    FLIM-FT &  &  & \\

    VGG &  &  & \\
    VGG-FT & $0.872 \pm 0.007$ & $0.844 \pm 0.015$ & $0.851 \pm 0.014 $ \\

  \hline
  \end{tabular}
  \end{center}
  \caption{Results.}
  \label{tab:results}
\end{table}

\begin{table*}
  \begin{center}
  \begin{tabular}{|l|c|c|c|c|c|c| }
  \hline
   Method & Pr0 & Pr1 & Re0 & Re1 & Fs0 & Fs1 \\
  \hline\hline
      FLIM + SVM & $0.933 \pm 0.003 $ & $ 0.612 \pm 0.038$ & $ 0.838 \pm 0.024$ & $0.809 \pm 0.007 $ & $ 0.883 \pm 0.014$ & $0.697 \pm 0.025 $ \\ 
       VGG-FE + SVM & $0.941 \pm 0.005$ & $0.580 \pm 0.013$ & $0.809 \pm 0.014 $ & $ 0.839 \pm 0.016$ & $ 0.870 \pm 0.006$ & $0.686 \pm 0.005 $ \\\hline

      FLIM & $0.929 \pm 0.002$ & $0.654 \pm 0.016$ & $0.869 \pm 0.010$ & $0.789 \pm 0.009 $ & $0.898 \pm 0.004 $ & $ 0.715 \pm 0.006$ \\
    FLIM-FT &  &  &  &  &  &  \\

    VGG &  &  &  &  &  & \\
    VGG-FT & $0.948 \pm 0.001$ & $0.629 \pm 0.032$ & $0.841 \pm 0.021 $ & $ 0.854 \pm 0.006$ & $ 0.891 \pm 0.012$ & $0.724 \pm 0.020 $  \\

  \hline
  \end{tabular}
  \end{center}
  \caption{Results per class.}
  \label{tab:results_per_class}
\end{table*}

We tested three scenarios: (I) using out method to learn the feature extractor weights and the classifier with backpropagation; (II) using our method as initialization and train the whole model with backpropagation; (III) using SVM Linear as classification. For VGG-16, we tested two scenarios: (IV) initializing with Xavier initialization and training from scratch; (V) using ImageNet weights and training with the training set images. The results can be seen in Table \ref{tab:results}. As the experiments were repeated three times, we reported the mean and standard deviation of each the metrics in the test set.
